\documentclass[12pt]{report}


\input{macros}

\begin{document}

\author{Edited by\\
  H.J. Lu\thanks{hongjiu.lu@intel.com},
  Michael Matz\thanks{matz@suse.de},
  Milind Girkar\thanks{milind.girkar@intel.com},
  Jan Hubi\v{c}ka\thanks{jh@suse.cz},\\
  Andreas Jaeger\thanks{aj@suse.de},
  Mark Mitchell\thanks{mark@codesourcery.com}}

\title{System V Application Binary Interface\\
{\Large AMD64 Architecture Processor Supplement\\
(With LP64 and ILP32 Programming Models)\\
Draft Version \version}}
\maketitle
\tableofcontents
\listoftables
\listoffigures

\section*{Revision History}

\begin{description}


\item[0.99] Add description of TLS relocations (thanks to Alexandre
  Oliva) and mention the decimal floating point and AVX types (thanks to
  H.J. Lu).
\item[0.98] Various clarifications and fixes according to feedback
  from Sun, thanks to Terrence Miller.  DWARF register numbers for some
  system registers, thanks to Jan Beulich.  Add \code{R_X86_64_SIZE32} and
  \code{R_X86_64_SIZE64} relocations; extend meaning of \code{e_phnum}
  to handle more than 0xffff program headers, thanks to Rod Evans.
  Add footnote about passing of \code{decimal} datatypes. 
  Specify that \code{_Bool} is booleanized at the caller.
\item[0.97] Integrate Fortran ABI.
\item[0.96] Use \code{SHF_X86_64_LARGE} instead \code{SHF_AMD64_LARGE}
  (thanks to Evandro Menezes).  Correct various grammatical errors
  noted by Mark F. Haigh, who also noted that there are no global VLAs
  in C99.  Thanks also to Robert R. Henry.
\item[0.95] Include description of the medium PIC memory model (thanks
  to Jan Hubi\v{c}ka) and large model (thanks to Evandro Menezes).
\item[0.94] Add sections in Development Environment, Program Loading,
  a description of EH_FRAME sections and general cleanups to make
  text in this ABI self-contained.  Thanks to Michael Walker and Terrence
  Miller.
\item[0.93] Add sections about program headers, new section types and
  special sections for unwinding information.  Thanks to Michael
  Walker.
\item[0.92] Fix some typos (thanks to Bryan Ford), add section about
  stack layout in the Linux kernel.  Fix example in figure~
  \ref{fig_passing_example} (thanks to Tom Horsley).  Add section on
  unwinding through assembler (written by Michal Ludvig).  Remove
  \code{mmxext} feature (thanks to Evandro Menezes).  Add section on
  Fortran (by Steven Bosscher) and stack unwinding (by Jan
  Hubi\v{c}ka).
\item[0.91] Clarify that x87 is default mode, not MMX (by Hans Peter
  Anvin).
\item[0.90]
  Change DWARF register
  numbers again; mention that \code{__m128} needs alignment; fix typo
  in figure~\ref{fig-stack-frame}; add some comments on kernel
  expectations; mention TLS extensions; add example for passing of
  variable-argument lists; change semantics of \RAX in
  variable-argument lists; improve formatting; mention that X87 class
  is not used for passing; make \code{/lib64} a Linux specific
  section; rename x86-64 to AMD64; describe passing of complex types.
  Special thanks to Andi Kleen, Michal Ludvig, Michael Matz, David
  O'Brien and Eric Young for their comments.
\item[0.21] Define \code{__int128} as class INTEGER in register
  passing.  Mention that \reg{al} is used for variadic argument lists.  Fix
  some textual problems.  Thanks to H. Peter Anvin, Bo Thorsen, and
  Michael Matz.
\item[0.20 --- 2002-07-11] Change DWARF register number values of
  \RBX, \RSI, \RSI (thanks to Michal Ludvig).  Fix footnotes for
  fundamental types (thanks to H. Peter Anvin). Specify \code{size\_t}
  (thanks to Bo Thorsen and Andreas Schwab).  Add new section on
  floating point environment functions.
\item[0.19 --- 2002-03-27] Set name of Linux dynamic linker, mention
  \reg{fs}.
  Incorporate changes from H. Peter Anvin <hpa@zytor.com>
  for booleans and define handling of sub-64-bit integer types in
  registers.
\end{description}

\include{introduction}
\include{sw-installation}
\include{low-level-sys-info}
\include{object-files}
\include{dl}
\include{libraries}
\include{development}
\include{execution}

%%% Bah, we can't nest includes, hence begin the chapter outside
%%% of the included file, so that conventions and fortran come to lie
%%% in the same chapter
\chapter{Conventions}\editornote{This chapter is used to document some features special to
  the \xARCH ABI.  The different sections might be moved to another
  place or removed completely.}
\include{conventions}
\include{fortran}
\include{x32}
\include{secure}
\include{mpx}

\appendix
\include{kernel}
\include{linker-optimization}
\include{abi-index}

\end{document}
