%%% vim:ai:tw=72:
\chapter{Low Level System Information}

\section{Machine Interface}

\subsection{Processor Architecture}

\subsection{Data Representation}

Within this specification, the term \emph{\textindex{\byte{}}} refers to
a 8-bit object, the term \emph{\textindex{\twobyte{}}} refers to a 16-bit
object, the term \emph{\textindex{\fourbyte{}}} refers to a 32-bit
object, the term \emph{\textindex{\eightbyte{}}} refers to a 64-bit
object, and the term \emph{\textindex{\sixteenbyte{}}} refers to a
128-bit object.%
\footnote{The \intelabi uses the term \emph{\textindex{halfword}} for
  a 16-bit object, the term \emph{\textindex{word}} for a 32-bit
  object, the term \emph{\textindex{doubleword}} for a 64-bit object.  But
  most IA-32 processor specific documentation define a
  \emph{\textindex{word}} as a 16-bit object, a
  \emph{\textindex{doubleword}} as a 32-bit object, a
  \emph{\textindex{quadword}} as a 64-bit object and a
  \emph{\textindex{double quadword}} as a 128-bit object.}

\subsubsection{Fundamental Types}

Figure~\ref{basic-types} shows the correspondence between ISO C's
scalar types and the processor's.  \code{__int128}, \code{__float128},
\code{__m64}, \code{__m128} and \code{__m256} types are optional.

\begin{figure}
  \caption{Scalar Types}\label{basic-types}
{ % Use small here - the table is still too large
  % Has anybody an idea how to shrink the table so that it fits the page?
  \small
  \begin{tabular}{l|l|c|c|l}
    \hline\noalign{\smallskip}
     & &  & \multicolumn{1}{c|}{Alignment} & \multicolumn{1}{c|}{\xARCH} \\
    \multicolumn{1}{c|}{Type} & \multicolumn{1}{c|}{C}
     &  \texttt{sizeof} & (bytes)
     & \multicolumn{1}{c|}{Architecture}  \\
    \hline
    & \texttt{_Bool}$^\dagger$ & 1 & 1 & boolean \\
    \cline{2-5}
    & \texttt{char}        & 1 & 1 & signed byte \\
    & \texttt{signed char} & & \\
    \cline{2-5}
    & \texttt{unsigned char} & 1 & 1 & unsigned byte \\
    \cline{2-5}
    & \texttt{short} & 2 & 2 & signed \twobyte \\
    & \texttt{signed short} & & \\
    \cline{2-5}
    & \texttt{unsigned short} & 2 & 2 & unsigned \twobyte \\
    \cline{2-5}
    & \texttt{int} & 4 & 4 & signed \fourbyte \\
    Integral & \texttt{signed int} & & \\
    & \texttt{enum}$^{\dagger\dagger\dagger}$ & & \\
    \cline{2-5}
    & \texttt{unsigned int} & 4 & 4 & unsigned \fourbyte \\
    \cline{2-5}
    & \texttt{long} & 8 & 8 & signed \eightbyte \\
    & \texttt{signed long} & & \\
    & \texttt{long long} & & \\
    & \texttt{signed long long} & & \\
    \cline{2-5}
    & \texttt{unsigned long} & 8 & 8 & unsigned \eightbyte \\
    & \texttt{unsigned long long} & 8 & 8 & unsigned \eightbyte \\
    \cline{2-5}
    & \texttt{__int128}$^{\dagger\dagger}$ & 16 & 16 & signed \sixteenbyte \\
    & \texttt{signed __int128}$^{\dagger\dagger}$ & 16 & 16 & signed \sixteenbyte \\
    \cline{2-5}
    & \texttt{unsigned __int128}$^{\dagger\dagger}$ & 16 & 16 & unsigned \sixteenbyte \\
    \hline
    Pointer & \texttt{\textit{any-type} *} & 8 & 8 & unsigned \eightbyte \\
    & \texttt{\textit{any-type} (*)()} & & \\
    \hline
    Floating-& \texttt{float} & 4 & 4 & single (IEEE-754) \\
    \cline{2-5}
    point & \texttt{double} & 8 & 8 & double (IEEE-754) \\
    \cline{2-5}
    & \texttt{long double} & 16 & 16 & 80-bit extended (IEEE-754) \\
    \cline{2-5}
    & \texttt{__float128}$^{\dagger\dagger}$ & 16 & 16 & 128-bit extended (IEEE-754) \\
    \hline
    Decimal-& \texttt{_Decimal32} & 4 & 4 & 32bit BID (IEEE-754R) \\
    \cline{2-5}
    floating-& \texttt{_Decimal64} & 8 & 8 & 64bit BID (IEEE-754R) \\
    \cline{2-5}
    point & \texttt{_Decimal128} & 16 & 16 & 128bit BID (IEEE-754R) \\
    \hline
    Packed & \texttt{__m64}$^{\dagger\dagger}$ & 8 & 8 & \MMX{} and \threednow \\
    \cline{2-5}
    & \texttt{__m128}$^{\dagger\dagger}$ & 16 & 16 & SSE and SSE-2 \\
    \cline{2-5}
    & \texttt{__m256}$^{\dagger\dagger}$ & 32 & 32 & AVX \\
\noalign{\smallskip}
\cline{1-5}
\multicolumn{3}{l}{\small $^\dagger$ This type is called \texttt{bool}
in C++.}\\
\multicolumn{3}{l}{\small $^{\dagger\dagger}$ These types are optional.}\\
\multicolumn{5}{p{14cm}}{\small $^{\dagger\dagger\dagger}$ C++ and some
implementations of C permit enums larger than an int.  The underlying
type is bumped to an unsigned int, long int or unsigned long int, in
that order.}\\
  \end{tabular}
}
\end{figure}

The \codeindex{__float128} type uses a 15-bit exponent, a 113-bit
mantissa (the high order significant bit is implicit) and an exponent
bias of 16383.\footnote{Initial implementations of the \xARCH
  architecture are expected to support operations on the
  \texttt{__float128} type only via software emulation.}

The \code{long double} type uses a 15 bit exponent, a 64-bit mantissa
with an explicit high order significant bit and an exponent bias of
16383.\footnote{This type is the x87 double extended precision data
  type.}  Although a \code{long double} requires 16 bytes of storage,
only the first 10 bytes are significant.  The remaining six bytes are
tail padding, and the contents of these bytes are undefined.

The \code{__int128} type is stored in little-endian order in memory,
i.e., the 64 low-order bits are stored at a a lower address than the
64 high-order bits.

A null pointer (for all types) has the value zero.

The type \codeindex{size\_t} is defined as \code{unsigned long}.

Booleans\index{boolean}, when stored in a memory object, are stored as
single byte objects the value of which is always 0 (\code{false}) or 1
(\code{true}).  When stored in integer registers (except for passing
as arguments), all 8 bytes of the register are significant;
any nonzero value is considered \code{true}.

Like the Intel386 architecture, the \xARCH architecture in general
does not require all data accesses to be properly aligned.  Misaligned
data accesses are slower than aligned accesses
but otherwise behave identically.  The only exceptions are that
\code{__m128} and \code{__m256} must always be aligned properly.

\subsubsection{Aggregates and Unions}

Structures and unions assume the alignment of their most strictly
aligned component.  Each member is assigned to the lowest available
offset with the appropriate alignment.  The size of any object is always
a multiple of the object`s alignment.

An array uses the same alignment as its elements, except that a local
or global array variable of length at least 16 bytes or a C99
variable-length array variable always has alignment
of at least 16 bytes.\footnote{The alignment requirement allows the
  use of SSE instructions when operating on the array.  The compiler
  cannot in general calculate the size of a variable-length array (VLA), but
  it is expected that most VLAs will require at least 16 bytes, so it
  is logical to mandate that VLAs have at least a 16-byte alignment.}

Structure and union objects can require padding to meet size and
alignment constraints.  The contents of any padding is undefined.

%%% SUN <examples - add diagrams>

\subsubsection{Bit-Fields}

C struct and union definitions may include bit-fields that define
integral values of a specified size.

%%% SUN <table><include complete i386 + below table>
\begin{figure}[h]
\Hrule
  \caption{Bit-Field Ranges}
  \begin{center}
    \leavevmode
    \begin{tabular}{l|l|l}
      \multicolumn{1}{c}{Bit-field Type}
         & \multicolumn{1}{c}{Width $w$}
         & \multicolumn{1}{c}{Range} \\
      \hline
      \texttt{signed char} & & $-2^{w - 1}$ to $2^{w-1}-1$ \\
      \texttt{char} & 1 to 8 & 0 to $2^{w}-1$ \\
      \texttt{unsigned char} & & 0 to $2^{w}-1$ \\
      \hline
      \texttt{signed short} & & $-2^{w - 1}$ to $2^{w-1}-1$ \\
      \texttt{short} & 1 to 16 & 0 to $2^{w}-1$ \\
      \texttt{unsigned short} & & 0 to $2^{w}-1$ \\
      \hline
      \texttt{signed int} & & $-2^{w - 1}$ to $2^{w-1}-1$ \\
      \texttt{int} & 1 to 32 & 0 to $2^{w}-1$ \\
      \texttt{unsigned int} & & 0 to $2^{w}-1$ \\
      \hline
      \texttt{signed long} & & $-2^{w - 1}$ to $2^{w-1}-1$ \\
      \texttt{long} & 1 to 64 & 0 to $2^{w}-1$ \\
      \texttt{unsigned long} & & 0 to $2^{w}-1$ \\
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

The ABI does not permit bit-fields having the type \texttt{__m64},
\texttt{__m128} or \texttt{__m256}.  Programs using bit-fields of
these types are not portable.

Bit-fields that are neither signed nor unsigned
always have non-negative values. Although they may have type char,
short, int, or long (which can have negative values), these bit-fields
have the same range as a bit-field of the same size with the
corresponding unsigned type. Bit-fields obey the same size and
alignment rules as other structure and union members.

Also:
\begin{itemize}
 \item bit-fields are allocated from right to left
 \item bit-fields must be contained in a storage unit appropriate 
       for its declared type
 \item bit-fields may share a storage unit with other struct / union members
\end{itemize}

Unnamed bit-fields' types do not affect the alignment of a structure
or union.

%%% SUN <examples - add diagrams>

\section{Function Calling Sequence}

This section describes the standard function calling sequence,
including stack frame layout, register usage, parameter passing and so
on.

The standard calling sequence requirements apply only to global
functions.  Local functions that are not reachable from other
compilation units may use different conventions.  Nevertheless, it is
recommended that all functions use the standard calling sequence when
possible.

\subsection{Registers and the Stack Frame}
\label{subsec-registers}

The \xARCH architecture provides 16 general purpose 64-bit registers.
In addition the architecture provides 16 SSE registers, each 128 bits
wide and 8 x87 floating point registers, each 80 bits wide.  Each of
the x87 floating point registers may be referred to in \MMX/\threednow
mode as a 64-bit register.  All of these registers are global to all
procedures active for a given thread.

Intel AVX (Advanced Vector Extensions) provides 16 256-bit wide AVX registers
(\reg{ymm0} - \reg{ymm15}).  The lower 128-bits of \reg{ymm0} - \reg{ymm15}
are aliased to the respective 128b-bit SSE registers (\reg{xmm0} -
\reg{xmm15}). For purposes of parameter passing and function return,
\reg{xmmN} and \reg{ymmN} refer to the same register. Only one of them
can be used at the same time.  We use vector register to refer to either
SSE or AVX register.

This subsection discusses usage of each register.  Registers \RBP, \RBX and
\reg{r12} through \reg{r15} ``belong'' to the calling function and the
called function is required to preserve their values.  In other words,
a called function must preserve these registers' values for its
caller.  Remaining registers ``belong'' to the called
function.\footnote{Note that in contrast to the \intelabi, \RDI,
  and \RSI belong to the called function, not the caller.}  If a
calling function wants to preserve such a register value across a
function call, it must save the value in its local stack frame.

The CPU shall be in x87 mode upon entry to a function.  Therefore,
every function that uses the \MMX registers is required to issue an
\op{emms} or \op{femms} instruction after using \MMX registers, before
returning or calling another function.  \footnote{All x87 registers
are caller-saved, so callees that make use of the \MMX registers may
use the faster \op{femms} instruction.}  The direction flag \code{DF} in the
\reg{rFLAGS} register must be clear (set to ``forward'' direction) on function
entry and return.  Other user flags have no specified role in the
standard calling sequence and are {\em not} preserved across calls.

The control bits of the \code{MXCSR} register are callee-saved
(preserved across calls), while the status bits are caller-saved (not
preserved).  The x87 status word register is caller-saved, whereas
the x87 control word is callee-saved.

\subsection{The Stack Frame}
\label{sec-stack-frame}

In addition to registers, each function has a frame on the run-time
stack.  This stack grows downwards from high addresses.  Figure
\ref{fig-stack-frame} shows the stack organization.

\begin{figure}
\Hrule
  \caption{Stack Frame with Base Pointer}
  \label{fig-stack-frame}
  \begin{center}
    \begin{tabular}{r|c|l}
      \noalign{\smallskip}
      \multicolumn{1}{l}{Position} &
      \multicolumn{1}{c}{Contents} &
      \multicolumn{1}{l}{Frame} \\
      \noalign{\smallskip}  \cline{1-3}
      \code{8n+16(\RBP)} & memory argument \eightbyte $n$ \\
      & \dots & Previous \\
      \code{16(\RBP)} & memory argument \eightbyte $0$ \\
      \cline{1-3}
      \code{8(\RBP)} & return address \\ \cline{2-2}
      \code{0(\RBP)} & previous \RBP value \\
      \cline{2-2}
      \code{-8(\RBP)} & unspecified & Current \\
      & \dots & \\
      \code{0(\RSP)} & variable size \\
      \cline{2-2}
      \code{-128(\RSP)} & red zone\\
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

The end of the input argument area shall be aligned on a 16 (32, if
\texttt{__m256} is passed on stack) byte boundary.  In other
words, the value $(\RSP + 8)$ is always a multiple of $16$ ($32$) when
control is transferred to the function entry point.  The
stack pointer, \RSP, always points to the end of the latest allocated
stack frame.  \footnote{The conventional use of \RBP{} as a frame
  pointer for the stack frame may be avoided by using \RSP (the stack
  pointer) to index into the stack frame.  This technique saves two
  instructions in the prologue and epilogue and makes one additional
  general-purpose register (\RBP) available.}

The 128-byte area beyond the location pointed to by \RSP is considered
to be reserved and shall not be modified by signal or interrupt
handlers.\footnote{Locations within 128 bytes can be addressed using
  one-byte displacements.}  Therefore, functions may use this area for
temporary data that is not needed across function calls.  In
particular, leaf functions may use this area for their entire stack
frame, rather than adjusting the stack pointer in the prologue and
epilogue.  This area is known as the \textindex{red zone}.

\subsection{Parameter Passing}
\label{sec-calling-conventions}

After the argument values have been computed, they are placed either in
registers or pushed on the stack.  The way how values are passed is
described in the following sections.

\paragraph{Definitions}
We first define a number of classes to classify arguments.  The
classes are corresponding to \xARCH register classes and defined as:

\begin{description}
\item[INTEGER] This class consists of integral types that fit into one of
  the general purpose registers.
\item[SSE] The class consists of types that fit into a vector register.
\item[SSEUP] The class consists of types that fit into a vector
  register and can be passed and returned in the upper bytes of it.
\item[X87, X87UP] These classes consists of types that will be returned via
  the x87 FPU.
\item[COMPLEX\_X87] This class consists of types that will be returned
  via the x87 FPU.
\item[NO\_CLASS] This class is used as initializer in the algorithms.  It
  will be used for padding and empty structures and unions.
\item[MEMORY] This class consists of types that will be passed and
  returned in memory via the stack.
\end{description}


\paragraph{Classification}
The size of each argument gets rounded up to
\eightbytes.\footnote{Therefore the stack will always be \eightbyte aligned.}

The basic types are assigned their natural classes:
\begin{itemize}
\item Arguments of types (signed and unsigned) \code{_Bool}, \code{char},
  \code{short}, \code{int}, \code{long}, \code{long long}, and
  pointers are in the INTEGER class.
\item Arguments of types \code{float}, \code{double}, \code{_Decimal32},
  \code{_Decimal64} and \code{__m64} are in class SSE.
\item Arguments of types \code{__float128}, \code{_Decimal128}
  and \code{__m128} are split into two halves.  The least significant
  ones belong to class SSE, the most significant one to class SSEUP.
\item Arguments of type \code{__m256} are split into four \eightbyte
  chunks.  The least significant one belongs to class SSE and all the
  others to class SSEUP.
\item The 64-bit mantissa of arguments of type \code{long double}
  belongs to class X87, the 16-bit exponent plus 6 bytes of padding
  belongs to class X87UP.
\item Arguments of type \code{__int128} offer the same operations as
INTEGERs, yet they do not fit into one general purpose register but
require two registers.  For
classification purposes \code{__int128} is treated as if it
were implemented as:
\begin{verbatim}
typedef struct {
  long low, high;
} __int128;
\end{verbatim}
with the exception that arguments of type \code{__int128} that are
stored in memory must be aligned on a 16-byte boundary.

\item Arguments of \code{complex T} where \code{T} is one of the types
  \code{float or double} are treated as if they are
  implemented as:
\begin{verbatim}
struct complexT {
  T real;
  T imag;
};
\end{verbatim}
\item A variable of type \code{complex long double} is classified as
  type COMPLEX\_X87.
\end{itemize}


The classification of aggregate (structures and arrays) and union
types works as follows:

\begin{enumerate}
\item If the size of an object is larger than four \eightbytes, or
  it contains unaligned fields, it has class MEMORY
  \footnote{The post merger clean up described later ensures that,
  for the processors that do not support the \code{__m256} type, if
  the size of an object is larger than two \eightbytes and the first
  \eightbyte is not SSE or any other \eightbyte is not SSEUP, it still
  has class MEMORY. This in turn ensures that for processors that
  do support the \code{__m256} type, if the size of an object is
  four \eightbytes and the first eightbyte is SSE and all other
  eightbytes are SSEUP, it can be passed in a register.}.

\item If a C++ object has either a non-trivial copy constructor
    or a non-trivial destructor
  \footnote{A de/constructor is trivial if it is an implicitly-declared
             default de/constructor and if:
    \begin{itemize}
        \item its class has no virtual functions
          and no virtual base classes, and

        \item all the direct base classes of its class have trivial
          de/constructors, and

        \item for all the nonstatic data members of its class that are
          of class type (or array thereof), each such class has a
          trivial de/constructor.
    \end{itemize}},
    it is passed by invisible reference
    (the object is replaced in the parameter list by a pointer that
    has class INTEGER)
  \footnote{An object with either a non-trivial copy
   constructor or a non-trivial destructor cannot be passed by value
   because such objects must have well defined addresses.  Similar
   issues apply when returning an object from a function.}.

\item If the size of the aggregate exceeds a single \eightbyte, each is
    classified separately.  Each \eightbyte gets initialized to class NO_CLASS.

\item Each field of an object is classified recursively so that always
   two fields are considered.  The resulting class is calculated
   according to the classes of the fields in the \eightbyte:
   \begin{enumerate}
   \item
      If both classes are equal, this is the resulting class.
   \item If one of the classes is NO_CLASS, the resulting class is the other class.
   \item If one of the classes is MEMORY, the result is the MEMORY class.
   \item If one of the classes is INTEGER, the result is the INTEGER.
   \item If one of the classes is X87, X87UP, COMPLEX\_X87 class,
     MEMORY is used as class.
   \item Otherwise class SSE is used.
   \end{enumerate}
\item Then a post merger cleanup is done:
  \begin{enumerate}
  \item If one of the classes is MEMORY, the whole argument is passed in memory.
  \item If X87UP is not preceded by X87, the whole argument is passed in
    memory.
  \item If the size of the aggregate exceeds two \eightbytes and the first
    \eightbyte isn't SSE or any other \eightbyte isn't SSEUP, the whole
    argument is passed in memory.
  \item If SSEUP is not preceded by SSE or SSEUP, it is converted to SSE.
  \end{enumerate}
\end{enumerate}

\paragraph{Passing}
Once arguments are classified, the registers get assigned (in
left-to-right order) for passing as follows:

\begin{enumerate}
\item If the class is MEMORY, pass the argument on the stack.

\item If the class is INTEGER, the next available register of the
  sequence \RDI, \RSI, \RDX, \RCX, \reg{r8} and \reg{r9} is
  used\footnote{Note that \reg{r11} is neither required to be
    preserved, nor is it used to pass arguments.  Making this register
    available as scratch register means that code in the PLT need not
    spill any registers when computing the address to which control
    needs to be transferred.  \RAX is used to indicate the number of
    vector arguments passed to a function requiring a variable number of
    arguments. \reg{r10} is used for passing a function's static chain
    pointer.}.

\item If the class is SSE, the next available vector register is used, the
   registers are taken in the order from \reg{xmm0} to \reg{xmm7}.

\item If the class is SSEUP, the \eightbyte is passed in the next
   available \eightbyte chunk of the last used vector register.

\item If the class is X87, X87UP or COMPLEX\_X87, it is passed in memory.
\end{enumerate}

When a value of type \code{_Bool} is returned or passed in a register or
on the stack, bit 0 contains the truth value and bits 1 to 7 shall be
zero\footnote{Other bits are left unspecified, hence the consumer
side of those values can rely on it being 0 or 1 when truncated to 8
bit.}.

\begin{figure}
\Hrule
  \caption{Register Usage}
  \label{fig-reg-usage}
  \begin{center}
    \begin{tabular}{l|p{8.35cm}|l}
      \noalign{\smallskip}
      \multicolumn{1}{c}{} &
      \multicolumn{1}{c}{}&
      \multicolumn{1}{l}{Preserved across}\\
      \multicolumn{1}{c}{Register} &
      \multicolumn{1}{c}{Usage}&
      \multicolumn{1}{l}{function calls}\\
      \hline
\RAX & temporary register; with variable arguments passes
information about the number of vector registers used; 1$^{\rm st}$
return register & No \\
\RBX & callee-saved register; optionally used as base pointer & Yes \\
\RCX & used to pass 4$^{\rm th}$ integer argument to functions & No \\
\RDX & used to pass 3$^{\rm rd}$ argument to functions; 2$^{\rm nd}$ return register & No \\
\RSP & stack pointer & Yes \\
\RBP & callee-saved register; optionally used as frame pointer & Yes \\
\RSI & used to pass 2$^{\rm nd}$  argument to functions & No \\
\RDI & used to pass 1$^{\rm st}$  argument to functions & No \\
\reg{r8} & used to pass 5$^{\rm th}$  argument to functions & No \\
\reg{r9} & used to pass 6$^{\rm th}$  argument to functions & No \\
\reg{r10} & temporary register, used for passing a function's static
chain pointer & No \\
\reg{r11} & temporary register & No\\
\reg{r12--r15} & callee-saved registers & Yes \\
\reg{xmm0}--\reg{xmm1} & used to pass and return floating point
arguments & No\\
\reg{xmm2}--\reg{xmm7} & used to pass floating point arguments & No\\
\reg{xmm8}--\reg{xmm15} & temporary registers & No\\
\reg{mmx0}--\reg{mmx7}& temporary registers & No\\
\reg{st0},\reg{st1} & temporary registers; used to return \code{long double} arguments & No \\
\reg{st2}--\reg{st7} & temporary registers & No \\
\reg{fs}& Reserved for system (as thread specific data register) & No\\
\code{mxcsr}& SSE2 control and status word & partial\\
\code{x87 SW}& x87 status word & No\\
\code{x87 CW}& x87 control word & Yes\\
    \end{tabular}

  \end{center}
\Hrule
\end{figure}

If there are no registers available for any \eightbyte of an
argument, the whole argument is passed on the stack. If registers have
already been assigned for some \eightbytes of such an argument, the
assignments get reverted.

Once registers are assigned, the arguments passed in memory are pushed
on the stack in reversed (right-to-left\footnote{Right-to-left order
  on the stack makes the handling of functions that take a variable
  number of arguments simpler.  The location of the first argument can
  always be computed statically, based on the type of that argument.
  It would be difficult to compute the address of the first argument
  if the arguments were pushed in left-to-right order.}) order.

For calls that may call functions that use varargs or stdargs
(prototype-less calls or calls to functions containing ellipsis
(\dots) in the declaration) \reg{al}%
\footnote{Note that the rest of \RAX is undefined, only the contents
  of \reg{al} is defined.}
 is used as hidden argument to specify the number of vector registers
 used. The contents of \reg{al} do not need to
match exactly the number of registers, but must be an upper bound on
the number of vector registers used and is in the range 0--8 inclusive.

When passing \texttt{__m256} arguments to functions that use varargs
or stdarg, function prototypes must be provided.  Otherwise, the
run-time behavior is undefined.

\paragraph{Returning of Values}
The returning of values is done according to the following algorithm:
\begin{enumerate}
\item Classify the return type with the classification algorithm.

\item If the type has class MEMORY, then the caller provides space for
   the return value and passes the address of this storage in \RDI as
   if it were the first argument to the function.  In effect, this
   address becomes a ``hidden'' first argument.  This storage must
   not overlap any data visible to the callee through other names
   than this argument.

   On return \RAX will contain the address that has been passed in by
   the caller in \RDI.

\item If the class is INTEGER, the next available register of the
   sequence \RAX, \RDX is used.

\item If the class is SSE, the next available vector register of the
   sequence \reg{xmm0}, \reg{xmm1} is used.

\item If the class is SSEUP, the \eightbyte is returned in the next
   available \eightbyte chunk of the last used vector register.

\item If the class is X87, the value is returned on the X87 stack in
   \reg{st0} as 80-bit x87 number.

\item If the class is X87UP, the value is returned together with the
   previous X87 value in \reg{st0}.
\item If the class is COMPLEX\_X87, the real part of the value is
  returned in \reg{st0} and the imaginary part in \reg{st1}.
\end{enumerate}

As an example of the register passing conventions, consider the
declarations and the function call shown in
Figure \ref{fig_passing_example}.  The corresponding register
allocation is given in Figure \ref{fig_allocation_example}, the stack
frame offset given shows the frame before calling the function.

\begin{figure}[H]
\Hrule
\caption{Parameter Passing Example}
\label{fig_passing_example}
\begin{center}
\code{
\begin{tabular}{|l|}
\cline{1-1}
typedef struct \{ \\
\ \ int a, b;\\
\ \ double d;\\
\} structparm;\\
structparm s;\\
int e, f, g, h, i, j, k;\\
long double ld;\\
double m, n;\\
__m256 y;\\
\\
extern void func (int e, int f,\\
\phantom{extern void func (}structparm s, int g, int h,\\
\phantom{extern void func (}long double ld, double m,\\
\phantom{extern void func (}__m256 y,\\
\phantom{extern void func (}double n, int i, int j, int k);\\
\\
func (e, f, s, g, h, ld, m, y, n, i, j, k);\\
\cline{1-1}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Register Allocation Example}
\label{fig_allocation_example}
\begin{center}
\begin{tabular}{ll|ll|ll}
\multicolumn{2}{c}{General Purpose Registers} &
\multicolumn{2}{c}{Floating Point Registers} &
\multicolumn{2}{c}{Stack Frame Offset}\\
\hline
\RDI:    &\code{e}      &\reg{xmm0}:&\code{s.d}&\code{0:} &\code{ld} \\
\RSI:    &\code{f}      &\reg{xmm1}:&\code{m}  &\code{16:}&\code{j} \\
\RDX:    &\code{s.a,s.b}&\reg{ymm2}:&\code{y}  &\code{24:}&\code{k} \\
\RCX:    &\code{g}      &\reg{xmm3}:&\code{n}  &          & \\
\reg{r8}:&\code{h}      &           &          &          & \\
\reg{r9}:&\code{i}      &           &          &          & \\
\end{tabular}

\end{center}
\Hrule
\end{figure}

\section{Operating System Interface}

\subsection{Exception Interface}

As the \xARCH manuals describe, the processor changes mode to handle
\emph{\textindex{exceptions},} which may be synchronous,
floating-point/coprocessor or asynchronous.
Synchronous and floating-point/coprocessor exceptions,
being caused by instruction execution, can be explicitly generated
by a process. This section, therefore, specifies those exception types
with defined behavior. The \xARCH architecture classifies exceptions as
\emph{faults}, \emph{traps}, and \emph{aborts}. See the \intelabi
for more information about their differences.

\subsubsection{Hardware Exception Types}
The operating system defines the correspondence between hardware
exceptions and the signals specified by \codeindex{signal}(BA\_OS)
as shown in table \ref{tab-hw-exceptions}. Contrary to the i386
architecture, the \xARCH does not define any instructions that
generate a bounds check fault in long mode.

\begin{table}
\Hrule
  \caption{Hardware Exceptions and Signals}
  \label{tab-hw-exceptions}
  \begin{center}
    \begin{tabular}[t]{c|l|l}
      \multicolumn{1}{c}{Number} & \multicolumn{1}{c}{Exception name}
         & \multicolumn{1}{c}{Signal}\\
      \hline
      \texttt{0} & \texttt{divide error fault} & \texttt{SIGFPE} \\
      \texttt{1} & \texttt{single step trap/fault} & \texttt{SIGTRAP} \\
      \texttt{2} & \texttt{non-maskable interrupt} & \texttt{none} \\
      \texttt{3} & \texttt{breakpoint trap} & \texttt{SIGTRAP} \\
      \texttt{4} & \texttt{overflow trap} & \texttt{SIGSEGV} \\
      \texttt{5} & \texttt{(reserved)} &\\
      \texttt{6} & \texttt{invalid opcode fault} & \texttt{SIGILL} \\
      \texttt{7} & \texttt{no coprocessor fault} & \texttt{SIGFPE} \\
      \texttt{8} & \texttt{double fault abort} & \texttt{none} \\
      \texttt{9} & \texttt{coprocessor overrun abort} & \texttt{SIGSEGV} \\
      \texttt{10} & \texttt{invalid TSS fault} & \texttt{none} \\
      \texttt{11} & \texttt{segment no present fault} & \texttt{none} \\
      \texttt{12} & \texttt{stack exception fault} & \texttt{SIGSEGV} \\
      \texttt{13} & \texttt{general protection fault/abort}&\texttt{SIGSEGV} \\
      \texttt{14} & \texttt{page fault} & \texttt{SIGSEGV} \\
      \texttt{15} & \texttt{(reserved)} &\\
      \texttt{16} & \texttt{coprocessor error fault} & \texttt{SIGFPE} \\
      \texttt{other} & \texttt{(unspecified)} & \texttt{SIGILL}
    \end{tabular}
  \end{center}
\Hrule
\end{table}

\begin{table}
\Hrule
  \caption{Floating-Point Exceptions}
  \begin{center}
    \begin{tabular}[t]{l|l}
      \multicolumn{1}{c}{Code} & \multicolumn{1}{c}{Reason} \\
      \hline
      \texttt{FPE\_FLTDIV} & \texttt{floating-point divide by zero} \\
      \texttt{FPE\_FLTOVF} & \texttt{floating-point overflow} \\
      \texttt{FPE\_FLTUND} & \texttt{floating-point underflow} \\
      \texttt{FPE\_FLTRES} & \texttt{floating-point inexact result} \\
      \texttt{FPE\_FLTINV} & \texttt{invalid floating-point operation}
    \end{tabular}
  \end{center}
\Hrule
\end{table}

\subsection{Virtual Address Space}

Although the \xARCH architecture uses 64-bit pointers, implementations
are only required to handle 48-bit addresses.  Therefore, conforming
processes may only use addresses from \texttt{0x00000000\,00000000} to
\texttt{0x00007fff\,ffffffff}\footnote{0x0000ffff\,ffffffff is not a
  canonical address and cannot be used.}.

Processes begin with three logical segments, commonly called text,
data, and stack. Use of shared libraries add other segments and
a process may dynamically create segments.

\subsection{Page Size}

Systems are permitted to use any power-of-two page size between 4KB
and 64KB, inclusive.

\subsection{Virtual Address Assignments}

Conceptually processes have the full address space available.
In practice, however, several factors limit the size of a process.
\begin{itemize}
  \item The system reserves a configuration dependent amount of virtual space.
  \item The system reserves a configuration dependent amount of space per
    process.
  \item
    A process whose size exceeds the system's available combined physical
    memory and secondary storage cannot run. Although some physical memory
    must be present to run any process, the system can execute processes that
    are bigger than physical memory, paging them to and from secondary storage.
    Nonetheless, both physical memory and secondary storage are
    shared resources. System load, which can vary from one program execution
    to the next, affects the available amount.
\end{itemize}

Programs that dereference null pointers are erroneous and a process
should not expect 0x0 to be a valid address.

\begin{figure}[H]
\Hrule
  \caption{Virtual Address Configuration}
  \label{fig-address}
  \begin{center}
    \begin{tabular}{r|c|l}
      \noalign{\smallskip}  \cline{2-2}
      \verb|0xffffffffffffffff| & Reserved system area & End of memory\\
      & \dots & \\ \cline{2-2}
      & \dots & \\
      \verb|0x80000000000| & Dynamic segments & \\ \cline{2-2}
      & \dots & \\
      \verb|0| & Process segments & Beginning of memory\\ \cline{2-2}
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

Although applications may control their memory assignments, the typical
arrangement appears in figure \ref{fig-cfg}.

\begin{figure}[H]
\Hrule
  \caption{Conventional Segment Arrangements}
  \label{fig-cfg}
  \begin{center}
    \begin{tabular}{r|c|l}
      \cline{2-2}
      & \dots & \\
      \verb|0x80000000000| & Dynamic segments & \\ \cline{2-2}
      & Stack segment & \\
      & \dots & \\ \cline{2-2}
      & \dots & \\
      & Data segments & \\ \cline{2-2}
      & \dots & \\
      \verb|0x400000| & Text segments & \\ \cline{2-2}
      \verb|0| & Unmapped & \\ \cline{2-2}
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

\section{Process Initialization}

\subsection{Initial Stack and Register State}

\subsubsection{Special Registers}

The \xARCH architecture defines floating point instructions.  At
process startup the two floating point units, SSE2 and x87, both have
all floating-point exception status flags cleared.  The status of the
control words is as defined in tables \ref{x87-fpucw} and
\ref{mxcsr-status}.

\begin{table}[H]
\Hrule
  \caption{x87 Floating-Point Control Word}
  \label{x87-fpucw}
  \begin{center}
    \begin{tabular}[t]{l|l|l}
      \multicolumn{1}{c}{Field} & \multicolumn{1}{c}{Value}& \multicolumn{1}{c}{Note} \\
      \hline
      \texttt{RC} & 0 & Round to nearest\\
      \texttt{PC} & 11& Double extended precision\\
      \texttt{PM} & 1 & Precision masked\\
      \texttt{UM} & 1 & Underflow masked\\
      \texttt{OM} & 1 & Overflow masked\\
      \texttt{ZM} & 1 & Zero divide masked\\
      \texttt{DM} & 1 & De-normal operand masked\\
      \texttt{IM} & 1 & Invalid operation masked\\
    \end{tabular}
  \end{center}
\Hrule
\end{table}

\begin{table}[H]
\Hrule
  \caption{MXCSR Status Bits}
  \label{mxcsr-status}
  \begin{center}
    \begin{tabular}[t]{l|l|l}
      \multicolumn{1}{c}{Field} & \multicolumn{1}{c}{Value}& \multicolumn{1}{c}{Note} \\
      \hline
      \texttt{FZ}  & 0 & Do not flush to zero\\
      \texttt{RC}  & 0 & Round to nearest\\
      \texttt{PM}  & 1 & Precision masked\\
      \texttt{UM}  & 1 & Underflow masked\\
      \texttt{OM}  & 1 & Overflow masked\\
      \texttt{ZM}  & 1 & Zero divide masked\\
      \texttt{DM}  & 1 & De-normal operand masked\\
      \texttt{IM}  & 1 & Invalid operation masked\\
      \texttt{DAZ} & 0 & De-normals are not zero\\
    \end{tabular}
  \end{center}
\Hrule
\end{table}

The \code{rFLAGS} register contains the system flags, such as the
direction flag and the carry flag.  The low 16 bits (FLAGS portion)
of \code{rFLAGS} are accessible by
application software.  The state of them at process initialization
is shown in table \ref{rflags-status}.

\begin{table}[H]
\Hrule
  \caption{\code{rFLAGS} Bits}
  \label{rflags-status}
  \begin{center}
    \begin{tabular}[t]{l|l|l}
      \multicolumn{1}{c}{Field} & \multicolumn{1}{c}{Value}& \multicolumn{1}{c}{Note} \\
      \hline
      \texttt{DF} & 0 & Direction forward\\
      \texttt{CF} & 0 & No carry\\
      \texttt{PF} & 0 & Even parity\\
      \texttt{AF} & 0 & No auxiliary carry\\
      \texttt{ZF} & 0 & No zero result\\
      \texttt{SF} & 0 & Unsigned result\\
      \texttt{OF} & 0 & No overflow occurred\\
    \end{tabular}
  \end{center}
\Hrule
\end{table}

\subsubsection{Stack State}

This section describes the machine state that \codeindex{exec}(BA\_OS) creates
for new processes. Various language implementations transform this
initial program state to the state required by the language standard.

For example, a C program begins executing at a
function named \code{main} declared as:

\begin{footnotesize}
\begin{verbatim}
    extern int main ( int argc , char *argv[ ] , char* envp[ ] );
\end{verbatim}
\end{footnotesize}

where
\begin{description}
 \item[argc] is a non-negative argument count
 \item[argv] is an array of argument strings, with \code{argv[argc] == 0}
 \item[envp] is an array of environment strings,
             terminated by a null pointer.
\end{description}

When \code{main()} returns its value is passed to \code{exit()} and if
that has been
over-ridden and returns, \code{\_exit()} (which must be immune to user
interposition).

The initial state of the process stack, i.e. when \code{\_start} is called
is shown in figure \ref{initial-stack}. 

\begin{figure}[H]
\Hrule
\caption{Initial Process Stack}
\label{initial-stack}
\begin{center}
\begin{tabular}{p{14em}|l|p{7em}}
  \multicolumn{1}{c}{Purpose}
         & \multicolumn{1}{c}{Start Address}
         & \multicolumn{1}{c}{Length} \\
\hline
  Unspecified & High Addresses & \\ \hline
  Information block, including argument strings, environment strings,
    auxiliary information ... & & varies \\ \hline
  Unspecified & & \\ \hline
  Null auxiliary vector entry & & 1 eightbyte \\ \hline
  Auxiliary vector entries ... & & 2 eightbytes each \\ \hline
  0 & & eightbyte \\ \hline
  Environment pointers ... & & 1 eightbyte each \\ \hline
  0 & \code{8+8*argc+\RSP} & eightbyte \\ \hline
  Argument pointers & \code{8+\RSP} & argc eightbytes \\ \hline
  Argument count & \RSP & eightbyte \\ \hline
  Undefined & Low Addresses & \\ \hline
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

Argument strings, environment strings, and the auxiliary information
appear in no specific order within the information block and they
need not be compactly allocated.

Only the registers listed below have specified values at process
entry:
\begin{description}
 \item[\RBP] The content of this register is unspecified at process
                initialization time, but the user code should mark the
                deepest stack frame by setting the frame pointer to zero.
 \item[\RSP] The stack pointer holds the address of the byte with
                lowest address which is part of the stack. It is
                guaranteed to be 16-byte aligned at process entry.
 \item[\RDX] a function pointer that the application should register
                with \code{atexit}(BA\_OS).
\end{description}

It is unspecified whether 
the data and stack segments are initially mapped with
execute permissions or not.
Applications which need to execute code on the stack or data
segments should take proper precautions, e.g., by calling
\code{mprotect()}.

\subsection{Thread State}

New threads inherit the floating-point state of the parent thread
and the state is private to the thread thereafter.

\subsection{Auxiliary Vector}

The \textindex{auxiliary vector} is an array of the following structures
(ref. figure \ref{fig_auxv_t}),
interpreted according to the \code{a\_type} member.

\begin{figure}[H]
\Hrule
\caption{\code{auxv\_t} Type Definition}
\label{fig_auxv_t}
\begin{center}
\begin{verbatim}
 typedef struct
 {
     int a_type;
     union {
         long a_val;
         void *a_ptr;
         void (*a_fnc)();
     } a_un;
 } auxv_t;
\end{verbatim}
\end{center}
\Hrule
\end{figure}

The \xARCH ABI uses the auxiliary vector types defined in figure \ref{aux-vec}.

\begin{figure}[H]
\Hrule
\caption{Auxiliary Vector Types}
\label{aux-vec}
\begin{center}
\begin{tabular}{l|r|l}
  \multicolumn{1}{c}{Name}
         & \multicolumn{1}{c}{Value}
         & \multicolumn{1}{c}{\code{a\_un}} \\
      \hline
\texttt{AT_NULL}& 0 & ignored\\
\texttt{AT_IGNORE}& 1& ignored\\
\texttt{AT_EXECFD}& 2& \texttt{a_val}\\
\texttt{AT_PHDR}& 3& \texttt{a_ptr}\\
\texttt{AT_PHENT}& 4& \texttt{a_val}\\
\texttt{AT_PHNUM}& 5& \texttt{a_val}\\
\texttt{AT_PAGESZ}& 6& \texttt{a_val}\\
\texttt{AT_BASE}& 7& \texttt{a_ptr}\\
\texttt{AT_FLAGS}& 8& \texttt{a_val}\\
\texttt{AT_ENTRY}& 9& \texttt{a_ptr}\\
\texttt{AT_NOTELF}& 10& \texttt{a_val}\\
\texttt{AT_UID}& 11& \texttt{a_val}\\
\texttt{AT_EUID}& 12& \texttt{a_val}\\
\texttt{AT_GID}& 13& \texttt{a_val}\\
\texttt{AT_EGID}& 14& \texttt{a_val}\\
\hline
    \end{tabular}
  \end{center}
\Hrule
\end{figure}

\begin{description}
\item[AT_NULL] The auxiliary vector has no fixed length; instead its
  last entry's \code{a\_type} member has this value.
\item[AT_IGNORE] This type indicates the entry has no meaning. The
  corresponding value of \code{a\_un} is undefined.
\item[AT_EXECFD] At process creation the system may pass control to an
  interpreter program. When this happens, the system places either an
  entry of type \code{AT\_EXECFD} or one of type \code{AT\_PHDR} in the
  auxiliary vector. The entry for type \code{AT\_EXECFD} uses the
  \code{a\_val} member to contain a file descriptor open to read the
  application program's object file.
\item[AT_PHDR] The system may create the memory image of the
  application program before passing control to the interpreter program.
  When this happens, the \code{a\_ptr} member of the \code{AT\_PHDR}
  entry tells the interpreter where to find the program header table in
  the memory image.
\item[AT_PHENT] The \code{a\_val} member of this entry holds the size,
  in bytes, of one entry in the program header table to which the
  \code{AT\_PHDR} entry points.
\item[AT_PHNUM] The \code{a\_val} member of this entry holds the number of
  entries in the program header table to which the \code{AT\_PHDR} entry
  points.
\item[AT_PAGESZ] If present, this entry's \code{a\_val} member gives the
  system page size, in bytes.
\item[AT_BASE] The \code{a\_ptr} member of this entry holds the base address
  at which the interpreter program was loaded into memory. See ``Program
  Header'' in the System V ABI for more information about the base
  address.
\item[AT_FLAGS] If present, the \code{a\_val} member of this entry holds
  one-bit flags.  Bits with undefined semantics are set to zero.
\item[AT_ENTRY] The \code{a\_ptr} member of this entry holds the entry point
  of the application program to which the interpreter program should
  transfer control.
\item[AT_NOTELF] The \code{a_val} member of this entry is non-zero if
  the program is in another format than ELF.
\item[AT_UID] The \code{a_val} member of this entry holds the real
  user id of the process.
\item[AT_EUID] The \code{a_val} member of this entry holds the
  effective user id of the process.
\item[AT_GID] The \code{a_val} member of this entry holds the
  real group id of the process.
\item[AT_EGID] The \code{a_val} member of this entry holds the
  effective group id of the process.
\end{description}


\section{Coding Examples}
\label{sec_coding_examples}

This section discusses example code sequences for fundamental
operations such as calling functions, accessing static objects, and
transferring control from one part of a program to another.  Unlike
previous material, this material is not normative.

\subsection{Architectural Constraints}

The \xARCH architecture usually does not allow an instruction to encode
arbitrary
64-bit constants as immediate operand.  Most
instructions accept 32-bit immediates that are sign extended to the
64-bit ones.  Additionally the 32-bit operations with register
destinations implicitly perform zero extension making loads of 64-bit
immediates with upper half set to 0 even cheaper.

Additionally the branch instructions accept 32-bit immediate operands
that are sign extended and used to adjust the instruction pointer.
Similarly an instruction pointer relative addressing mode exists for
data accesses with equivalent limitations.

In order to improve performance and reduce code size, it is desirable
to use different \textindex{code models} depending on the
requirements.

Code models define constraints for symbolic values that allow the
compiler to generate better code.  Basically code models differ in
addressing (absolute versus position independent), code size, data
size and address range.  We define only a small number of code models
that are of general interest:

\begin{description}
\item[\textindex{Small code model}]
  The virtual address of code executed is known at link time.
  Additionally all symbols are known to be located in the virtual
  addresses in the range from $0$ to $2^{31}-2^{24}-1$ or from $0x00000000$ to $0x7effffff$\footnote{
    The number $24$ is chosen arbitrarily.  It allows for all memory of
    objects of size up to $2^{24}$ or 16M bytes to be addressed directly because the
    base address of such objects is constrained to be less than $2^{31}-2^{24}$ or $0x7f000000$.
    Without such constraint only the base address would be accessible
    directly, but not any offsetted variant of it.}.

  This allows the compiler to encode symbolic references with offsets
  in the range from $-(2^{31})$ to $2^{24}$ or from $0x80000000$ to $0x01000000$ directly in the sign
  extended immediate operands, with offsets in the range from $0$ to
  $2^{31}-2^{24}$ or from $0x00000000$ to $0x7f000000$ in the zero extended immediate operands and use
  instruction pointer relative addressing for the symbols with offsets
  in the range $-(2^{24})$ to $2^{24}$ or $0xff000000$ to $0x01000000$.

  This is the fastest code model and we expect it to be suitable for
  the vast majority of programs.

\item[\textindex{Kernel code model}]

  The kernel of an operating system is usually rather small but runs
  in the negative half of the address space.  So we define all symbols
  to be in the range from $2^{64}-2^{31}$ to $2^{64}-2^{24}$ or from $0xffffffff80000000$ to $0xffffffffff000000$.

  This code model has advantages similar to those of the small model,
  but allows encoding of zero extended symbolic references only for
  offsets from $2^{31}$ to $2^{31}+2^{24}$ or from $0x80000000$ to $0x81000000$. The range offsets for
  sign extended reference changes to $0$ to $2^{31}+2^{24}$ or $0x00000000$ to $0x81000000$.

\item[\textindex{Medium code model}]

  In the medium model, the data section is split into two parts --- the
  data section still limited in the same way as in the small code model
  and the large data section having no limits except for available
  addressing space.  The program layout must be set in a way so that
  large data sections (\code{.ldata}, \code{.lrodata}, \code{.lbss})
  come after the text and data sections.

  This model requires the compiler to use \code{movabs} instructions
  to access large static data and to load addresses into registers, but
  keeps the advantages of the small code model for manipulation of
  addresses in the small data and text sections (specially needed for
  branches).

  By default only data larger than 65535 bytes will be placed in the
  large data section.

\item[\textindex{Large code model}]

  The large code model makes no assumptions about addresses and sizes
  of sections.

  The compiler is required to use the \code{movabs} instruction, as in
  the medium code model, even for dealing with addresses inside the
  text section.  Additionally, indirect branches are needed when
  branching to addresses whose offset from the current instruction
  pointer is unknown.

  It is possible to avoid the limitation on the text section in the
  small and medium models by breaking up the program into multiple
  shared libraries, so this model is strictly only required if the text
  of a single function becomes larger than what the medium model allows.

\item[\textindex{Small position independent code model} (\textindex{PIC})]

  Unlike the previous models, the virtual addresses of instructions
  and data are not known until dynamic link time.  So all addresses
  have to be relative to the instruction pointer.

  Additionally the maximum distance between a symbol and the end of an
  instruction is limited to $2^{31}-2^{24}-1$ or $0x7effffff$, allowing the compiler
  to use instruction pointer relative branches and addressing modes
  supported by the hardware for every symbol with an offset in the
  range $-(2^{24})$ to $2^{24}$ or $0xff000000$ to $0x01000000$.

\item[\textindex{Medium position independent code model}
  (\textindex{PIC})]

  This model is like the previous model, but similarly to the medium static
  model adds large data sections at the end of object files.

  In the medium PIC model, the instruction pointer relative addressing
  can not be used directly for accessing large static data, since the
  offset can exceed the limitations on the size of the displacement
  field in the instruction.  Instead an unwind sequence consisting of
  \code{movabs}, \code{lea} and \code{add} needs to be used.

\item[\textindex{Large position independent code model}
  (\textindex{PIC})]

  This model is like the previous model, but makes no assumptions
  about the distance of symbols.

  The large PIC model implies the same limitation as the medium
  PIC model regarding addressing of static data.  Additionally,
  references to the global offset table and to the procedure linkage
  table and branch destinations need to be calculated in a similar
  way.  Further the size of the text segment is allowed to be up to
  16EB in size, hence similar restrictions apply to all address
  references into the text segments, including branches.

\end{description}

\subsection{Conventions}

In this document some special assembler symbols are used in the coding
examples and discussion.  They are:

\begin{itemize}
\item \code{name@GOT}: specifies the offset to the GOT entry for
      the symbol \code{name} from the base of the GOT.
\item \code{name@GOTPLT}: specifies the offset to the GOT entry for
      the symbol \code{name} from the base of the GOT, implying that
      there is a corresponding PLT entry.
\item \code{name@GOTOFF}: specifies the offset to the location of
      the symbol \code{name} from the base of the GOT.
\item \code{name@GOTPCREL}: specifies the offset to the GOT entry
      for the symbol \code{name} from the current code location.
\item \code{name@PLT}: specifies the offset to the PLT entry of
      symbol \code{name} from the current code location.
\item \code{name@PLTOFF}: specifies the offset to the PLT entry of
      symbol \code{name} from the base of the GOT.
\item \code{_GLOBAL_OFFSET_TABLE_}: specifies the offset to the base
      of the GOT from the current code location.
\end{itemize}

\subsection{Position-Independent Function Prologue}

In the small code model all addresses (including GOT entries) are
accessible via the IP-relative addressing provided by the \xARCH
architecture.  Hence there is no need for an explicit GOT pointer and
therefore no function prologue for setting it up is necessary.

In the medium and large code models a register has to be allocated to hold the
address of the GOT in position-independent objects, because the \xARCH ISA does
not support an immediate displacement larger than 32 bits.

As \code{\%r15} is preserved across function calls, it is initialized
in the function prolog to hold the GOT address\footnote{If, at code
generation-time, it is determined that either no other functions are
called (leaf functions), the called functions addresses can be
resolved and are within 2GB, or no global data objects are referred
to, it is not necessary to store the GOT address in \code{\%r15} and
the prolog code that initializes it may be omitted.} for non-leaf
functions which call other functions through the PLT. Other functions
are free to use any other register. Throughout this document,
\code{\%r15} will be used in examples.

\begin{figure}[H]
\Hrule
\caption{Position-Independent Function Prolog Code} \label{large_pic_prolog}
medium model:
\begin{footnotesize}
\begin{verbatim}
   leaq     _GLOBAL_OFFSET_TABLE_(%rip),%r15 # GOTPC32 reloc
\end{verbatim}
\end{footnotesize}
large model:
\begin{footnotesize}
\begin{verbatim}
   pushq    %r15                          # save %r15
   leaq     1f(%rip),%r11                 # absolute %rip
1: movabs   $_GLOBAL_OFFSET_TABLE_,%r15   # offset to the GOT (R_X86_64_GOTPC64)
   leaq     (%r11,%r15),%r15              # absolute address of the GOT
\end{verbatim}
\end{footnotesize}
\Hrule
\end{figure}%$

For the medium model the GOT pointer is directly loaded, for the large
model
the absolute value of \code{\%rip} is added to the relative offset to the base
of the GOT in order to obtain its absolute address (see figure
\ref{large_pic_prolog}).

\subsection{Data Objects}

This section describes only objects with static storage.
Stack-resident objects are excluded since programs always compute
their virtual address relative to the stack or frame pointers.

Because only the \code{movabs} instruction uses 64-bit addresses
directly, depending on the code model either \RIP-relative addressing
or building addresses in registers and accessing the memory through
the register has to be used.

For absolute addresses \RIP-relative encoding can be used in the small
model.  In the medium model the \code{movabs} instruction has to be
used for accessing addresses.

Position-independent code cannot contain absolute address.  To access
a global symbol the address of the symbol has to be loaded from the
Global Offset Table.  The address of the entry in the GOT can be
obtained with a \RIP-relative instruction in the small model.

\subsubsection{Small models}

\begin{figure}[H]
\Hrule
\caption{Absolute Load and Store (Small Model)}
\begin{center}
\code{
\begin{tabular}{|l|c|ll|}
\cline{1-1}\cline{3-4}
extern int src[65536];&&.extern& src\\
extern int dst[65536];&&.extern& dst\\
extern int *ptr;&&.extern& ptr\\
static int lsrc[65536];&&.local& lsrc\\
		&&.comm& lsrc,262144,4\\
static int ldst[65536];&&.local& ldst\\
		&&.comm& ldst,262144,4\\
static int *lptr;&&.local& lptr\\
		&&.comm& lptr,8,8\\
&&.text&\\
dst[0] = src[0]; && movl& src(\%rip), \%eax\\
	   && movl& \%eax, dst(\%rip)\\
&&&\\
ptr = dst[0];&&movq& \$dst, ptr(\%rip)\\
&&&\\
\hbox{}*ptr = src[0];&&movq& ptr(\%rip),\%rax\\
           &&movl& src(\%rip),\%edx\\
           &&movl& \%edx, (\%rax)\\
&&&\\
ldst[0] = lsrc[0]; && movl& lsrc(\%rip), \%eax\\
	   && movl& \%eax, ldst(\%rip)\\
&&&\\
lptr = ldst;&&movq& \$dst, lptr(\%rip)\\
&&&\\
\hbox{}*lptr = lsrc[0];&&movq& lptr(\%rip),\%rax\\
           &&movl& lsrc(\%rip),\%edx\\
           &&movl& \%edx, (\%rax)\\
\cline{1-1}\cline{3-4}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Position-Independent Load and Store (Small PIC Model)}
\begin{center}
\code{
\begin{tabular}{|l|c|ll|}
\cline{1-1}\cline{3-4}
extern int src[65536];&&.extern& src\\
extern int dst[65536];&&.extern& dst\\
extern int *ptr;&&.extern& ptr\\
static int lsrc[65536];&&.local& lsrc\\
		&&.comm& lsrc,262144,4\\
static int ldst[65536];&&.local& ldst\\
		&&.comm& ldst,262144,4\\
static int *lptr;&&.local& lptr\\
		&&.comm& lptr,8,8\\
           &&.text&\\
dst[0] = src[0]; && movq& src@GOTPCREL(\%rip), \%rax\\
	   && movl& (\%rax), \%edx\\
           && movq& dst@GOTPCREL(\%rip), \%rax\\
           && movl& \%edx, (\%rax)\\
&&&\\
ptr = dst;&& movq& ptr@GOTPCREL(\%rip), \%rax\\
           && movq& dst@GOTPCREL(\%rip), \%rdx\\
           && movq& \%rdx, (\%rax)\\
&&&\\
\hbox{}*ptr = src[0];&& movq &ptr@GOTPCREL(\%rip),\%rax\\
           && movq &(\%rax), \%rdx\\
           && movq &src@GOTPCREL(\%rip), \%rax\\
           && movl &(\%rax), \%eax\\
           && movl &\%eax, (\%rdx)\\
&&&\\
ldst[0] = lsrc[0]; && movl& lsrc(\%rip), \%eax\\
	   && movl& \%eax, ldst(\%rip)\\
&&&\\
lptr = ldst;&&lea& ldst(\%rip),\%rdx\\
           &&movq& \%rdx, lptr(\%rip)\\
&&&\\
\hbox{}*lptr = lsrc[0];&&movq& lptr(\%rip),\%rax\\
           &&movl& lsrc(\%rip),\%edx\\
           &&movl& \%edx, (\%rax)\\
\cline{1-1}\cline{3-4}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\subsubsection{Medium models}

\begin{figure}[H]
\Hrule
\caption{Absolute Load and Store (Medium Model)}
\begin{center}
\code{
\begin{tabular}{|l|c|ll|}
\cline{1-1}\cline{3-4}
extern int src[65536];&&.extern& src\\
extern int dst[65536];&&.extern& dst\\
extern int *ptr;&&.extern& ptr\\
static int lsrc[65536];&&.local& lsrc\\
		&&.comm& lsrc,262144,4\footnote{linker is responsible for placing the symbol in large data section}\\
static int ldst[65536];&&.local& ldst\\
		&&.comm& ldst,262144,4\\
static int *lptr;&&.local& lptr\\
		&&.comm& lptr,8,8\\
&&.text&\\
dst[0] = src[0]; && movabsl& src, \%eax\\
	   && movabsl& \%eax, dst\\
&&&\\
ptr = dst;&&movabsq& \$dst,\%rdx\\
           &&movq& \%rdx, ptr\\
&&&\\
\hbox{}*ptr = src[0];&&movq& ptr(\%rip),\%rdx\\
           &&movabsl& src,\%eax\\
           &&movl& \%eax, (\%rdx)\\
&&&\\
ldst[0] = lsrc[0]; && movabsl& lsrc, \%eax\\
	   && movabsl& \%eax, ldst\\
&&&\\
lptr = ldst;&&movabsq& \$ldst,\%rdx\\
           &&movabsq& \%rdx, lptr\\
&&&\\
\hbox{}*lptr = lsrc[0];&&movq& lptr(\%rip),\%rdx\\
           &&movabsl& lsrc,\%eax\\
           &&movl& \%eax, (\%rdx)\\
\cline{1-1}\cline{3-4}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Position-Independent Load and Store (Medium PIC Model)}
\begin{center}
\code{
\begin{tabular}{|l|c|ll|}
\cline{1-1}\cline{3-4}
extern int src[65536];&&.extern& src\\
extern int dst[65536];&&.extern& dst\\
extern int *ptr;&&.extern& ptr\\
static int lsrc[65536];&&.local& lsrc\\
		&&.comm& lsrc,262144,4\\
static int ldst[65536];&&.local& ldst\\
		&&.comm& ldst,262144,4\\
static int *lptr;&&.local& lptr\\
		&&.comm& lptr,8,8\\
           &&.text&\\
dst[0] = src[0]; && movq& src@GOTPCREL(\%rip), \%rax\\
	   && movl& (\%rax), \%edx\\
           && movq& dst@GOTPCREL(\%rip), \%rax\\
           && movl& \%edx, (\%rax)\\
&&&\\
ptr = dst;&& movq& ptr@GOTPCREL(\%rip), \%rax\\
           && movq& dst@GOTPCREL(\%rip), \%rdx\\
           && movq& \%rdx, (\%rax)\\
&&&\\
\hbox{}*ptr = src[0];&& movq &ptr@GOTPCREL(\%rip),\%rax\\
           && movq &(\%rax), \%rdx\\
           && movq &src@GOTPCREL(\%rip), \%rax\\
           && movl &(\%rax), \%eax\\
           && movl &\%eax, (\%rdx)\\
\cline{1-1}\cline{3-4}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}
\begin{figure}[H]
\Hrule
\caption{Position-Independent Load and Store (Medium PIC Model), continued}
\begin{center}
\code{
\begin{tabular}{|l|c|ll|}
\cline{1-1}\cline{3-4}
ldst[0] = lsrc[0]; && movabsq& lsrc@GOTOFF64, \%rax\\
	   && movl& (\%rax,\%r15), \%eax\\
	   && movabsq& ldst@GOTOFF64, \%rdx\\
	   && movl& \%eax, (\%rdx,\%r15)\\
&&&\\
lptr = ldst;&& movabsq& ldst@GOTOFF64, \%rax\\
           &&addq& \%r15, \%rax\\
           &&movq& \%rax, lptr(\%rip)\\
&&&\\
\hbox{}*lptr = lsrc[0];&& movabsq& lsrc@GOTOFF64, \%rax\\
           &&movl& (\%rax,\%r15),\%eax\\
	   &&movq& lptr(\%rip),\%rdx\\
           &&movl& \%eax, (\%rdx)\\
\cline{1-1}\cline{3-4}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\subsubsection{Large Models}

Again, in order to access data at any position in the 64-bit
addressing space, it is necessary to calculate the address
explicitly\footnote{ If, at code generation-time, it is
determined that a referred to global data object address is
resolved within 2GB, the \code{\%rip-relative} addressing
mode can be used instead. See example in figure
\ref{large_rip_rel_addr_abs}.}, not unlike the medium code
model.

\begin{figure}[H]
\Hrule
\caption{Absolute Global Data Load and Store}\label{large_abs_data_load_store}
\begin{center}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# static int src;  # & \verb#  Lsrc: .long                # \\
\verb# static int dst;  # & \verb#  Ldst: .long                # \\
\verb# extern int *ptr; # & \verb#        .extern  ptr         # \\
\hline
\verb# dst = src;       # & \verb#        movabs  $Lsrc,%rax   ; R_X86_64_64 # \\
\verb#                  # & \verb#        movabs  $Ldst,%rdx   ; R_X86_64_64 # \\
\verb#                  # & \verb#        movl    (%rax),%ecx  # \\
\verb#                  # & \verb#        movl    %ecx,(%rdx)  # \\
\hline
\verb# ptr = &dst;      # & \verb#        movabs  $ptr,%rax    ; R_X86_64_64 # \\
\verb#                  # & \verb#        movabs  $Ldst,%rdx   ; R_X86_64_64 # \\
\verb#                  # & \verb#        movq    %rdx,(%rax)  # \\
\hline
\verb# *ptr = src;      # & \verb#        movabs  $Lsrc,%rax   ; R_X86_64_64 # \\
\verb#                  # & \verb#        movabs  $ptr,%rdx    ; R_X86_64_64 # \\
\verb#                  # & \verb#        movl    (%rax),%ecx  # \\
\verb#                  # & \verb#        movq    (%rdx),%rdx  # \\ 
\verb#                  # & \verb#        movl    %ecx,(%rdx)  # \\
\hline
\end{tabular}
\end{footnotesize}
\end{center}
\end{figure}
            
\begin{figure}[H]
\Hrule
\caption{Faster Absolute Global Data Load and Store}
\label{large_rip_rel_addr_abs}
\begin{center}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# *ptr = src; # & \verb#         movabs    $ptr,%rdx        ; R_X86_64_64 # \\
\verb#             # & \verb#         movl      Lsrc(%rip),%ecx  # \\
\verb#             # & \verb#         movq      (%rdx),%rdx      # \\
\verb#             # & \verb#         movl      %ecx,(%rdx)      # \\
\hline
\end{tabular} %$
\end{footnotesize}
\end{center}
\end{figure}

For position-independent code access to both static and external global
data assumes that the GOT address is stored in a dedicated register.
In these examples we assume it is in \code{\%r15}\footnote{If,
at code generation-time, it is determined that a referred to global data
object address is resolved within 2GB, the \code{\%rip-relative}
addressing mode can be used instead. See example in figure
\ref{large_rip_rel_addr_pic}.} (see Function Prologue):

\begin{figure}[H]
\Hrule
\caption{Position-Independent Global Data Load and Store}
\label{large_pic_data_load_store}
\begin{center}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb#static int src;  # & \verb# Lsrc: .long                      # \\
\verb#static int dst;  # & \verb# Ldst: .long                      # \\
\verb#extern int *ptr; # & \verb#       .extern  ptr               # \\
\hline 
\verb# dst = src;      # & \verb#   movabs   $Lsrc@GOTOFF,%rax ; R_X86_64_GOTOFF64 # \\
\verb#                 # & \verb#   movabs   $Ldst@GOTOFF,%rdx ; R_X86_64_GOTOFF64 # \\
\verb#                 # & \verb#   movl     (%rax,%r15),%ecx  # \\
\verb#                 # & \verb#   movl     %ecx,(%rdx,%r15)  # \\
\hline
\verb# ptr = &dst;     # & \verb#   movabs $ptr@GOT,%rax       ; R_X86_64_GOT64 # \\
\verb#                 # & \verb#   movabs $Ldst@GOTOFF,%rdx   ; R_X86_64_GOTOFF64 # \\
\verb#                 # & \verb#   movq   (%rax,%r15),%rax    # \\
\verb#                 # & \verb#   leaq   (%rdx,%r15),%rcx    # \\
\verb#                 # & \verb#   movq   %rcx,(%rax)         # \\
\hline
\verb# *ptr = src;     # & \verb#   movabs $Lsrc@GOTOFF,%rax   ; R_X86_64_GOTOFF64 # \\
\verb#                 # & \verb#   movabs $ptr@GOT,%rdx       ; R_X86_64_GOT64 # \\
\verb#                 # & \verb#   movl   (%rax,%r15),%ecx    # \\
\verb#                 # & \verb#   movq   (%rdx,%r15),%rdx    # \\
\verb#                 # & \verb#   movl   %ecx,(%rdx)         # \\
\hline
\end{tabular}
\end{footnotesize}
\end{center}
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Faster Position-Independent Global Data Load and Store}
\label{large_rip_rel_addr_pic}
\begin{center}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# *ptr = src; # & \verb#     movabs    $ptr@GOT,%rdx    ; R_X86_64_GOT64 # \\
\verb#             # & \verb#     movl      Lsrc(%rip),%ecx  # \\
\verb#             # & \verb#     movq      (%rdx,%r15),%rdx # \\
\verb#             # & \verb#     movl      %ecx,(%rdx)      # \\
\hline
\end{tabular} %$
\end{footnotesize}
\end{center}
\end{figure}

\subsection{Function Calls}

\subsubsection{Small and Medium Models}

\begin{figure}[H]
\Hrule
\caption{Position-Independent Direct Function Call (Small and Medium Model)}
\begin{center}
\code{
\begin{tabular}{|l|c|l|}
\cline{1-1}\cline{3-3}
extern void function ();  &&.globl function\\
function ();              &&call function@PLT\\
\cline{1-1}\cline{3-3}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Position-Independent Indirect Function Call}
\begin{center}
\code{
\begin{tabular}{|l|c|l|}
\cline{1-1}\cline{3-3}
extern void (*ptr) ();    &&.globl ptr, name\\
extern void name ();      && \\
ptr = name;               &&movq ptr@GOTPCREL(\%rip), \%rax  \\
                          &&movq name@GOTPCREL(\%rip), \%rdx  \\
                          &&movq \%rdx, (\%rax)  \\
                          &&  \\
(*ptr)();                 &&movq ptr@GOTPCREL(\%rip), \%rax  \\
                          &&call *(\%rax)  \\
\cline{1-1}\cline{3-3}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\subsubsection {Large models}

It cannot be assumed that a function is within 2GB in
general. Therefore, it is necessary to explicitly calculate the
desired address reaching the whole 64-bit address space.

\begin{figure}[H]
\Hrule
\caption{Absolute Direct and Indirect Function Call} \label{large_abs_dir_func_call}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\code{static void (*ptr) (void);} & \verb#Lptr: .quad                  # \\
\code{extern void foo (void);}    & \verb#      .globl foo             # \\
\code{static void bar (void);}    & \verb#Lbar: ...                    # \\
\hline
\code{foo ();}                    & \verb#       movabs    $foo,%r11   ; R_X86_64_64 # \\
                                  & \verb#       call      *%r11       # \\
\code{bar ();}                    & \verb#       movabs    $Lbar,%r11  ; R_X86_64_64 # \\
                                  & \verb#       call      *%r11       # \\
\hline
\code{ptr = foo;}                 & \verb#       movabs    $Lptr,%rax  ; R_X86_64_64 # \\
                                  & \verb#       movabs    $foo,%r11   ; R_X86_64_64 # \\
                                  & \verb#       movq      %r11,(%rax) # \\
\code{ptr = bar;}                 & \verb#       movabs    $Lbar,%r11  ; R_X86_64_64 # \\
                                  & \verb#       movq      %r11,(%rax) # \\
\hline
\code{(*ptr) ();}                 & \verb#       movabs    $Lptr,%r11  ; R_X86_64_64 # \\
                                  & \verb#       call      *(%r11)     # \\
\hline
\end{tabular}
\\
\end{footnotesize}
\end{figure}

And in the case of position-independent objects
\footnote{See subsection ``Implementation advice'' for some optimizations.}:

\begin{figure}[H]
\Hrule
\caption{Position-Independent Direct and Indirect Function Call}\label{large_pic_pic_ind_func_call}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\code{static void (*ptr) (void);} & \verb#Lptr: .quad                      # \\
\code{extern void foo (void);}    & \verb#      .globl foo                 # \\
\code{static void bar (void);}    & \verb#Lbar: ...                        # \\
\hline
\code{foo ();}                    & \verb#movabs   $foo@GOT,%r11     ; R_x86_64_GOTPLT64 # \\
                                  & \verb#call     *(%r11,%r15)      # \\
\code{bar ();}                    & \verb#movabs   $Lbar@GOTOFF,%r11 ; R_X86_64_GOTOFF64 # \\
                                  & \verb#leaq     (%r11,%r15),%r11  # \\
                                  & \verb#call     *%r11             # \\
\hline
\code{ptr = foo;}                 & \verb#movabs   $Lptr@GOTOFF,%rax ; R_X86_64_GOTOFF64 # \\
                                  & \verb#movabs   $foo@PLTOFF,%r11  ; R_X86_64_PLTOFF64 # \\
                                  & \verb#leaq     (%r11,%r15),%r11  # \\
                                  & \verb#movq     %r11,(%rax,%r15)  # \\
\code{ptr = bar;}                 & \verb#movabs   $Lbar@GOTOFF,%r11 ; R_X86_64_GOTOFF64 # \\
                                  & \verb#leaq     (%r11,%r15),%r11  # \\
                                  & \verb#movq     %r11,(%rax,%r15)  # \\
\hline
\code{(*ptr) ();}                 & \verb#movabs $Lptr@GOTOFF,%r11   ; R_X86_64_GOTOFF64 # \\
                                  & \verb#call     *(%r11,%r15)      # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}

\subsubsection{Implementation advice}

If, at code generation-time, certain conditions are determined, it's
possible to generate faster or smaller code sequences as the large model
normally requires.  When:

\begin{description}
  \item[(absolute) target of function call is within 2GB], a direct call
  or \RIP-relative addressing might be used:\\
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb#bar ();#    & \verb# call   Lbar              # \\
\hline
\verb#ptr = bar;# & \verb# movabs $Lptr,%rax         ; R_X86_64_64 # \\
                  & \verb# leaq   $Lbar(%rip),%r11 # \\
                  & \verb# movq   %r11,(%rax)      # \\
\hline
\end{tabular}
\end{footnotesize}

  \item[(PIC) the base of GOT is within 2GB] an indirect call to the GOT
  entry might be implemented like so:\\
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\code{foo ();}    & \verb# call   *(foo@GOT)  ; R_X86_64_GOTPCREL # \\
\hline
\end{tabular}
\end{footnotesize}

  \item[(PIC) the base of PLT is within 2GB], the PLT entry may be referred to
  relatively to \RIP:\\
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\code{ptr = foo;} & \verb# movabs $Lptr@GOTOFF,%rax   ; R_X86_64_GOTOFF64 # \\
                  & \verb# leaq   $foo@PLT(%rip),%r11 ; R_X86_64_PLT32 # \\
                  & \verb# movq   %r11,(%rax,%r15)    # \\
\hline
\end{tabular}
\end{footnotesize}

  \item[(PIC) target of function call is within 2GB] and is either not global
  or bound locally, a direct call to the symbol may be used or it may be
  referred to relatively to \RIP:\\
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\code{bar ();}     & \verb#call      Lbar               # \\
\hline
\code{ptr = bar;}  & \verb#movabs    $Lptr@GOTOFF,%rax  ; R_X86_64_GOTOFF64 # \\
                   & \verb#leaq      $Lbar(%rip),%r11   # \\
                   & \verb#movq      %r11,(%rax,%r15)   # \\
\hline
\end{tabular}
\end{footnotesize}

\end{description}

\subsection{Branching}

\subsubsection{Small and Medium Models}

As all labels are within 2GB no special care has to be taken when
implementing branches.  The full \xARCH ISA is usable.

\subsubsection{Large Models}

Because functions can be theoretically up to 16EB long, the
maximum 32-bit displacement of conditional and unconditional
branches in the \xARCH ISA are not enough to address the
branch target. Therefore, a branch target address is
calculated explicitly \footnote{If, at code generation-time,
it is determined that the target addresses are within 2GB,
alternatively, branch target addresses may be calculated
implicitly (see figure \ref{large_cal_tgt_addr})}. For absolute objects:

\begin{figure}[H]
\Hrule
\caption{Absolute Branching Code}\label{large_abs_branching}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# if (!a)     # & \verb#        testl     %eax,%eax   # \\
\verb# {           # & \verb#        jnz       1f          # \\
\verb#             # & \verb#        movabs    $2f,%r11    ; R_X86_64_64 # \\
\verb#             # & \verb#        jmpq      *%r11       # \\
\verb#        ...  # & \verb# 1:      ...                  # \\
\verb# }           # & \verb# 2:                           # \\
\hline
\verb# goto Label; # & \verb#        movabs    $Label,%r11 ; R_X86_64_64 # \\
\verb#             # & \verb#        jmpq      *%r11       # \\
\verb# ...         # & \verb#        ...                   # \\
\verb# Label:      # & \verb# Label:                       # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Implicit Calculation of Target Address}\label{large_cal_tgt_addr}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# if (!a)     # & \verb#        testl   %eax,%eax  # \\
\verb# {           # & \verb#        jz      2f         # \\
\verb#        ...  # & \verb# 1:     ...                # \\
\verb# }           # & \verb# 2:                        # \\
\hline
\verb# goto Label; # & \verb#        jmp     Label      # \\
\verb# ...         # & \verb#        ...                # \\
\verb# Label:      # & \verb# Label:                    # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}

For position-independent objects:

\begin{figure}[H]
\Hrule
\caption{Position-Independent Branching Code}\label{large_pic_branching}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# if (!a)      # & \verb#     testl      %eax,%eax          # \\
\verb# {            # & \verb#     jnz        1f                 # \\
\verb#              # & \verb#     movabs     $2f@GOTOFF,%r11    ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#     leaq       (%r11,%r15),%r11   # \\
\verb#              # & \verb#     jmpq       *%r11              # \\
\verb#              # & \verb# 1:  ...                           # \\
\verb#         ...  # & \verb# 2:                                # \\
\verb# }            # & \verb#                                   # \\
\hline
\verb# goto Label;  # & \verb#     movabs     $Label@GOTOFF,%r11 ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#     leaq       (%r11,%r15),%r11   # \\
\verb#              # & \verb#     jmpq       *%r11              # \\
\verb# ...          # & \verb#                                   # \\
\verb# ...          # & \verb#     Label:                        # \\
\verb# Label:       # & \verb#                                   # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}

For absolute objects, the implementation of the \code{switch} statement is:

\begin{figure}[H]
\Hrule
\caption{Absolute Switch Code}\label{large_abs_switch}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# switch (a)     # & \verb#            cmpl   $0,%eax           # \\
\verb# {              # & \verb#            jl     .Ldefault         # \\
\verb#                # & \verb#            cmpl   $2,%eax           # \\
\verb#                # & \verb#            jg     .Ldefault         # \\
\verb#                # & \verb#            movabs $.Ltable,%r11  ; R_X86_64_64 # \\
\verb#                # & \verb#            jmpq   *(%r11,%eax,8)    # \\
\verb#                # & \verb#            .section .lrodata,"aLM",@progbits,8 # \\
\verb#                # & \verb#           .align 8                  # \\ 
\verb#                # & \verb#  .Ltable: .quad .Lcase0          ; R_X86_64_64 # \\
\verb#                # & \verb#           .quad .Ldefault        ; R_X86_64_64 # \\
\verb#                # & \verb#           .quad .Lcase2          ; R_X86_64_64 # \\
\verb#                # & \verb#           .previous                 # \\
\verb#       default: # & \verb#  .Ldefault:                         # \\
\verb#        ...     # & \verb#      ...                            # \\
\verb#       case 0:  # & \verb#  .Lcase0:                           # \\
\verb#       ...      # & \verb#       ...                           # \\
\verb#       case 2:  # & \verb#  .Lcase2:                           # \\
\verb#       ...      # & \verb#       ...                           # \\
\verb# }              # & \verb#                                     # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}

When building position-independent objects, the \code{switch}
statement implementation changes to:

\begin{figure}[H]
\Hrule
\caption{Position-Independent Switch Code}\label{large_pic_switch}
\begin{footnotesize}
\begin{tabular}{|l|l|}
\hline
\verb# switch (a)   # & \verb#   cmpl     $0,%eax              # \\
\verb# {            # & \verb#   jl       .Ldefault            # \\
\verb#              # & \verb#   cmpl     $2,%eax              # \\
\verb#              # & \verb#   jg       .Ldefault            # \\
\verb#              # & \verb#   movabs   $.Ltable@GOTOFF,%r11 ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#   leaq     (%r11,%r15),%r11     # \\
\verb#              # & \verb#   movq     *(%r11,%eax,8),%r11  # \\
\verb#              # & \verb#   leaq     (%r11,%r15),%r11     # \\
\verb#              # & \verb#   jmpq     *%r11                # \\
\verb#              # & \verb#   .section .lrodata,"aLM",@progbits,8 # \\
\verb#              # & \verb#   .align 8                      # \\
\verb#              # & \verb#.Ltable: .quad .Lcase0@GOTOFF    ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#         .quad .Ldefault@GOTOFF  ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#         .quad .Lcase2@GOTOFF    ; R_X86_64_GOTOFF64 # \\
\verb#              # & \verb#         .previous               # \\
\verb#     default: # & \verb#.Ldefault:                       # \\
\verb#     ...      # & \verb#   ...                           # \\
\verb#     case 0:  # & \verb#.Lcase0:                         # \\
\verb#     ...      # & \verb#   ...                           # \\
\verb#     case 2:  # & \verb#.Lcase2:                         # \\
\verb#     ...      # & \verb#   ...                           # \\
\verb# }            # & \verb#                                 # \\
\hline
\end{tabular}
\end{footnotesize}
\end{figure}\footnote{The
jump-table is emitted in a different section so as to occupy cache lines
without instruction bytes, thus avoiding exclusive cache
subsystems to thrash.}

\subsection{Variable Argument Lists}

Some otherwise portable C programs depend on the argument passing
scheme, implicitly assuming that all arguments are passed on the
stack, and arguments appear in increasing order on the stack.
Programs that make these assumptions never have been portable, but
they have worked on many implementations. However, they do not work on
the \xARCH architecture because some arguments are passed in
registers.  Portable C programs must use the header file
\code{<stdarg.h>} in order to handle variable argument lists.

When a function taking variable-arguments is called, \RAX\index{\RAX}
must be set to the total number of floating point parameters passed to
the function in vector registers.%
\footnote{This implies that the only legal values for \RAX when
  calling a function with variable-argument lists are 0 to 8
  (inclusive).}
%%% XXX: Really only floating pointer parameters?
%%% XXX: Use %al or %rax?

When \texttt{__m256} is passed as variable-argument, it should always
be passed on stack. Only named \texttt{__m256} arguments may be passed
in register as specified in section \ref{sec-calling-conventions}.

\begin{figure}[H]
\Hrule
\caption{Parameter Passing Example with Variable-Argument List}
\label{fig_va_passing_example}
\begin{center}
\code{
\begin{tabular}{|l|}
\cline{1-1}
int a, b;\\
long double ld;\\
double m, n;\\
__m256 u, y;\\
\\
extern void func (int a, double m, __m256 u, ...);\\
\\
func (a, m, u, b, ld, y, n);\\
\cline{1-1}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}

\begin{figure}[H]
\Hrule
\caption{Register Allocation Example for Variable-Argument List}
\label{fig_va_allocation_example}
\begin{center}
\begin{tabular}{ll|ll|ll}
\multicolumn{2}{c}{General Purpose Registers} &
\multicolumn{2}{c}{Floating Point Registers} &
\multicolumn{2}{c}{Stack Frame Offset}\\
\hline
\RDI: &\code{a}&\reg{xmm0}:&\code{m}&\code{0:} &\code{ld} \\
\RSI: &\code{b}&\reg{ymm1}:&\code{u}&\code{32:}&\code{y} \\
\RAX: & 3      &\reg{xmm2}:&\code{n}& \\
\end{tabular}
\end{center}
\Hrule
\end{figure}


\subsubsection{The Register Save Area\index{register save area}}

The prologue of a function taking a variable argument list and known
to call the macro \code{va_start} is expected to save the argument
registers to the \emph{register save area}.  Each argument register
has a fixed offset in the register save area as defined in the figure
\ref{fig-reg-save}.

Only registers that might be used to pass arguments need to be saved.
Other registers are not accessed and can be used for other purposes.  If a
function is known to never accept arguments passed in
registers\footnote{This fact may be determined either by exploring
  types used by the \code{va_arg}
macro, or by the fact that the named
  arguments already are exhausted the argument registers entirely.},
the register save area may be omitted entirely.

The prologue should use \RAX to avoid unnecessarily saving XMM
registers.  This is especially important for integer only programs to
prevent the initialization of the XMM unit.


\begin{figure}[H]
\Hrule
\caption{Register Save Area}
\label{fig-reg-save}
\begin{center}
\begin{tabular}{l|r}
\multicolumn{1}{c}{Register}&\multicolumn{1}{c}{Offset}\\
\hline
\RDI & $0$ \\
\RSI & $8$ \\
\RDX & $16$ \\
\RCX & $24$ \\
\reg{r8} & $32$ \\
\reg{r9} & $40$ \\
\reg{xmm0} & $48$ \\
\reg{xmm1} & $64$ \\
\dots &  \\
\reg{xmm15} & $288$ \\
\end{tabular}
\end{center}
\Hrule
\end{figure}

\subsubsection{The \codeindex{va_list} Type}

The \code{va_list} type is an array containing a single element of one
structure
containing the necessary information to implement the \code{va_arg} macro. The C
definition of \code{va_list} type is given in figure \ref{fig-va_list}.

\begin{figure}[H]
\Hrule
\caption{\code{va_list} Type Declaration}
\label{fig-va_list}
\begin{center}
\code{
\begin{tabular}{|l|}
\cline{1-1}
typedef struct \{\\
\ \ \ unsigned int gp_offset;\\
\ \ \ unsigned int fp_offset;\\
\ \ \ void *overflow_arg_area;\\
\ \ \ void *reg_save_area;\\
  \} va_list[1];\\
\cline{1-1}
\end{tabular}
}
\end{center}
\Hrule
\end{figure}


\subsubsection{The \codeindex{va_start} Macro}

The \code{va_start} macro initializes the structure as follows:

\begin{description}
\item [reg_save_area]
The element points to the start of the register save area.
\item [overflow_arg_area] This pointer is used to fetch arguments
  passed on the stack.  It is initialized with the address of the
  first argument passed on the stack, if any, and then always updated
  to point to the start of the next argument on the stack.
\item [gp_offset] The element holds the offset in bytes from
  \code{reg_save_area} to the place where the next available general
  purpose argument register is saved.  In case all argument registers
  have been exhausted, it is set to the value 48 ($6*8$).
\item [fp_offset]
The element holds the offset in bytes from \code{reg_save_area} to the
place where the next available floating point
argument register is saved.  In case all argument registers have been exhausted,
it is set to the value 304 ($6*8+16*16$).
\end{description}

\subsubsection{The \codeindex{va_arg} Macro}

The algorithm for the generic \code{va_arg(l, type)} implementation is
defined as follows:

\begin{enumerate}
\item
Determine whether \code{type} may be passed in the registers.  If not go to step \ref{stack}.
\item
Compute \code{num_gp} to hold the number of general purpose registers needed to pass \code{type}
and \code{num_fp} to hold the number of floating point registers needed.
\item
Verify whether arguments fit into registers.
In the case:
$$\code{l->gp_offset} > 48 - \code{num_gp} * 8$$ or $$\code{l->fp_offset} > 304 - \code{num_fp} * 16$$
go to step \ref{stack}.
\item
  Fetch \code{type} from \code{l->reg_save_area} with an offset of
  \code{l->gp_offset} and/or \code{l->fp_offset}.  This may require
  copying to a temporary location in case the parameter is passed in
  different register classes or requires an alignment greater than 8 for
  general purpose registers and 16 for XMM registers.
\item
Set:
$$\code{l->gp_offset} = \code{l->gp_offset} + \code{num_gp} * 8$$
$$\code{l->fp_offset} = \code{l->fp_offset} + \code{num_fp} * 16.$$
\item
Return the fetched \code{type}.
\item
\label{stack}
Align \code{l->overflow_arg_area} upwards to a 16 byte boundary if
alignment needed by \code{type} exceeds 8 byte boundary.
\item
Fetch \code{type} from \code{l->overflow_arg_area}.
\item
Set \code{l->overflow_arg_area} to:
$$\code{l->overflow_arg_area} + \code{sizeof} (\code{type})$$
\item
Align \code{l->overflow_arg_area} upwards to an 8 byte boundary.
\item
Return the fetched \code{type}.
\end{enumerate}

The \code{va_arg} macro is usually implemented as a compiler builtin and expanded in
simplified forms for each particular type.  Figure \ref{fig-va_arg} is a sample
implementation of the \code{va_arg} macro.
\begin{figure}[H]
\Hrule
\caption{Sample Implementation of \code{va_arg(l, int)}}
\label{fig-va_arg}
\begin{center}
\begin{tabular}{|llll|}
\cline{1-4}
&movl&\code{l->gp_offset}, \reg{eax}&\\
&cmpl&\$48, \reg{eax}&Is register available?\\
&jae&stack&If not, use stack\\
&leal&\$8(\RAX), \reg{edx}&Next available register\\
&addq&\code{l->reg_save_area}, \RAX&Address of saved register\\
&movl&\reg{edx}, \code{l->gp_offset}&Update \code{gp_offset}\\
&jmp&fetch&\\
stack:&movq&\code{l->overflow_arg_area}, \RAX&Address of stack slot\\
&leaq&8(\RAX), \RDX&Next available stack slot\\
&movq&\RDX,\code{l->overflow_arg_area}&Update\\
fetch:&movl&(\RAX), \reg{eax}&Load argument\\
\cline{1-4}
\end{tabular}
\end{center}
\Hrule
\end{figure}

\section{DWARF Definition}

This section\footnote{This section is structured in a way similar to
  the PowerPC psABI}
defines the Debug With Arbitrary Record Format (DWARF) debugging
format for the \xARCH processor family. The \xARCH ABI does not define
a debug format.  However, all systems that do implement DWARF on \xARCH shall
use the following definitions.

DWARF is a specification developed for symbolic, source-level debugging.
The debugging information format does not favor the design of any
compiler or debugger.  For more information on DWARF,
see \emph{DWARF Debugging Information Format},
revision: Version 3, January, 2006, Free Standards Group,
DWARF Standard Committee.  It's available at:
\url{http://www.dwarfstd.org/}.

\subsection{DWARF Release Number}

The DWARF definition requires some machine-specific definitions.
The register number mapping needs to be specified for the \xARCH
registers. In addition, the DWARF Version 3 specification
requires processor-specific address class codes to be defined.

\subsection{DWARF Register Number Mapping}

Table \ref{tbl-reg-num-map}\footnote{The table defines Return Address
  to have a register number, even though the address is stored in
  0(\RSP) and not in a physical register.}  outlines the register
number mapping for the \xARCH processor family.%
\footnote{This document does not define mappings for privileged registers.}%

\begin{figure}
\Hrule
\caption{DWARF Register Number Mapping} \label{tbl-reg-num-map}
\begin{center}
\begin{tabular}{l|r|l}
\multicolumn{1}{c}{Register Name}&\multicolumn{1}{c}{Number}&\multicolumn{1}{c}{Abbreviation}\\
\hline
General Purpose Register RAX & 0 &\RAX\\
General Purpose Register RDX & 1 &\RDX\\
General Purpose Register RCX & 2 &\RCX\\
General Purpose Register RBX & 3 &\RBX\\
General Purpose Register RSI & 4 &\RSI\\
General Purpose Register RDI & 5 &\RDI\\
Frame Pointer Register   RBP & 6 &\RBP\\
Stack Pointer Register   RSP & 7 &\RSP\\
Extended Integer Registers 8-15 & 8-15 &\reg{r8}--\reg{r15}\\
Return Address RA
& 16&\\
Vector Registers 0--7              & 17-24 & \reg{xmm0}--\reg{xmm7} \\
Extended Vector Registers 8--15    & 25-32 & \reg{xmm8}--\reg{xmm15} \\
Floating Point Registers 0--7   & 33-40 & \reg{st0}--\reg{st7} \\
MMX Registers 0--7              & 41-48 & \reg{mm0}--\reg{mm7} \\
Flag Register                   & 49    & \reg{rFLAGS} \\
Segment Register ES             & 50    & \reg{es} \\
Segment Register CS             & 51    & \reg{cs} \\
Segment Register SS             & 52    & \reg{ss} \\
Segment Register DS             & 53    & \reg{ds} \\
Segment Register FS             & 54    & \reg{fs} \\
Segment Register GS             & 55    & \reg{gs} \\
Reserved                        & 56-57 & \\
FS Base address                 & 58    & \reg{fs.base} \\
GS Base address                 & 59    & \reg{gs.base} \\
Reserved                        & 60-61 & \\
Task Register                   & 62    & \reg{tr} \\
LDT Register                    & 63    & \reg{ldtr} \\
128-bit Media Control and Status & 64   & \reg{mxcsr} \\
x87 Control Word                & 65    & \reg{fcw} \\
x87 Status Word                 & 66    & \reg{fsw} \\
\end{tabular}
\end{center}
\Hrule
\end{figure}

\section{Stack Unwind Algorithm}

The stack frames are not self descriptive and where stack unwinding is
desirable (such as for exception handling) additional unwind
information needs to be generated.  The information is stored in an
allocatable section \code{.eh_frame} whose format is identical to
\code{.debug_frame} defined by the DWARF debug information standard,
see \emph{DWARF Debugging Information Format}, with the following
extensions:

\begin{description}
\item[Position independence]

  In order to avoid load time relocations for position independent
  code, the FDE CIE offset pointer should be stored relative to the
  start of CIE table entry.  Frames using this extension of the DWARF
  standard must set the CIE identifier tag to 1.

\item[Outgoing arguments area delta]
  
  To maintain the size of the temporarily allocated outgoing arguments
  area present on the end of the stack (when using \code{push}
  instructions), operation \code{GNU_ARGS_SIZE} (\code{0x2e}) can be
  used.  This operation takes a single \code{uleb128} argument
  specifying the current size.  This information is used to adjust the
  stack frame when jumping into the exception handler of the function
  after unwinding the stack frame.  Additionally the CIE Augmentation
  shall contain an exact specification of the encoding used.  It is
  recommended to use a PC relative encoding whenever possible and
  adjust the size according to the code model used.

\item[CIE Augmentations:]
 
  The augmentation field is formated according to the augmentation
  field formating string stored in the CIE header.
  
  The string may contain the following characters:

\begin{description}
\item[z] Indicates that a \code{uleb128} is present determining the size of
  the augmentation section.
\item[L] Indicates the encoding (and thus presence) of an LSDA pointer in the FDE augmentation.
  
  The data filed consist of single byte specifying the way pointers
  are encoded.  It is a mask of the values specified by the table
  \ref{tbl-pointer-encoding}.

\begin{figure}
\Hrule
\caption{Pointer Encoding Specification Byte} \label{tbl-pointer-encoding}
\begin{center}
\begin{tabular}{r|l}
\multicolumn{1}{c}{Mask}&\multicolumn{1}{c}{Meaning}\\
\hline
0x1 & Values are stored as \code{uleb128} or \code{sleb128} type (according to flag 0x8)\\
0x2 & Values are stored as 2 bytes wide integers (\code{udata2} or \code{sdata2})\\
0x3 & Values are stored as 4 bytes wide integers (\code{udata4} or \code{sdata4})\\
0x4 & Values are stored as 8 bytes wide integers (\code{udata8} or \code{sdata8})\\
0x8 & Values are signed\\
0x10 & Values are PC relative\\
0x20 & Values are text section relative\\
0x30 & Values are data section relative\\
0x40 & Values are relative to the start of function\\
\end{tabular}
\end{center}
\Hrule
\end{figure}

The default DWARF3 pointer encoding (direct 4-byte absolute pointers)
is represented by value 0.

\item[R]  
  Indicates a non-default pointer encoding for FDE code pointers.  The
  formating is represented by a single byte in the same way as in the
  `L' command.

\item[P]  
  Indicates the presence and an encoding of a language personality
  routine in the CIE augmentation.  The encoding is represented by a
  single byte in the same way as in the 'L' command followed by a
  pointer to the personality function encoded by the specified
  encoding.

\end{description}

When the augmentation is present, the first command must always be
`\code{z}' to allow easy skipping of the information.

\end{description}

In order to simplify manipulation of the unwind tables, the runtime
library provide higher level API to stack unwinding mechanism, for
details see section \ref{unwindlib}.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "abi"
%%% End:
